Relationship between solar wind, $D_{st}$, and plasmasphere mass density on one-hour time scales

### 2015-09-03 paper.tex
New bibfile name for consistency; Makefile fixes

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period and 1991 and found that the average ion mass density, $M=\rho/n_e$, correlated with geomagnetic activity, with lower $D_{st}$ values corresponding to slightly average ion mass densities.  They also found a relationship between the average ion mass and 1.5- and 3-day averages of $K_p$ were compared to $M$.  Most measurements occurred in the MLT range of 12:00 and 18:00 in the plasma trough (in equatorial plane?)

\cite{Takahashi2010} developed a mass density dataset using measurements from the GOES satellites over the years 1980 through 1992, with most measurements in the range $L=6.8\pm0.2$ The equatorial mass density, $\rho_{eq}$ was estimated from the estimated mass density a power law dependence on the geocentric radius $R$, $\rho=\rho_{eq}(LR_E/R)$. A key result was that when daily-averaged measurements were considered, an enhancment in $\rho_{eq}$ appeared, indicating the possibility of a solar wind control over $\rho_{eq}$.

\cite{Takahashi2010} noted that downward spikes in the $D_{st}$ index coincide with with significant changes in $\rho_{eq}$ at an L-shell of 6.8~$R_E$. For five storms over a 20 day period, two had $\rho_{eq}$ spikes after the $D_{st}$ drop, two had $\rho_{eq}$ spikes before the drop, and one showed little change in $\rho_{eq}$. They then show an epoch analysis where $\rho_{eq}$ is seen to spike the day of a $D_{st}$ drop, using a daily average of 30 minute $\rho_{eq}$ and one hour $D_{st}$ measurements.

The results of \cite{Takahashi2006} and \cite{Takahashi2010} are consistent with other results on changes in mass density in the plasma trough region.  \cite{Hamilton1988} ... \cite{Gallagher1988} .... \cite{Yao2008} studied the relationship between $D_{st}$ and ion number density for different ions in different regions (ring current and plasma sheet) and found a general correlation during each of the four storms considered.

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset associated with \cite{Denton}, with data available from 1980-1991; all other parameters are from \cite{Reconstruction} over the time range of 1972-2013, which are on a 1-hour time grid. $\rho_{eq}$ is the inferred mass density based on the 3rd harmonic frequency of magnetic field measurements.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hourly range was used.  Fill values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple Geostationary Operational Environmental Satellites (GOES) satellites at the same time in the \cite{Denton} dataset, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data is given in Figure~\ref{AllData}.

@@

\newpage
\footnotesize
\bibliographystyle{plainnat}
\bibliography{reportbib}

@@

\begin{figure}[htp!]
\centering
\subcaptionbox{\label{ccplot:a}}
{\includegraphics[scale=0.42]{paperfigures/F107MDAllData.eps}}
\subcaptionbox{\label{ccplot:b}}
{\includegraphics[scale=0.42]{paperfigures/ccplot.eps}}
\caption{(a) 27-day averages of $F_{10.7}$ and $\rho_{eq}$. (b) Correlation between $log(\rho_{eq})$ and $F_{10.7}$ using hour, day, and 27-day averaging intervals; compare to \cite{Takahashi2010} Fig.~14.}
\label{ccplot}
\end{figure}
\clearpage

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/stormavs-dst-50-tak.eps}
\rule[1ex]{5cm}{0.5pt}
\includegraphics[scale=0.45]{paperfigures/stormavs-mass-tak.eps}
\caption{Events in the interval 1989-1991. Upper panel: $D_{st}$ events. Lower panel: $\rho_{eq}$ events.}
\label{DailyAverages}
\end{figure}

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/stormavs-dst-day.eps}
\rule[1ex]{5cm}{0.5pt}
\includegraphics[scale=0.45]{paperfigures/stormavs-mass-day.eps}
\caption{Events in the interval 1983-1991.  Upper panel: $D_{st}$ events. Lower panel: $\rho_{eq}$ events.}
\label{DailyAveragesAllData}
\end{figure}
\clearpage

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/stormavs-dst.eps}
\rule[1ex]{5cm}{0.5pt}
\includegraphics[scale=0.45]{paperfigures/stormavs-mass.eps}
\caption{(a) Average of solar wind and near-Earth measurements around the time $D_{st}$ crossed below $-50$~nT onset. (b) Same as (a) except around time intervals where mass density crossed above $40$~amu/cm$^3$.}
\label{Storms}
\end{figure}

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/stormavs-dd12.eps}
\rule[1ex]{5cm}{0.5pt}
\includegraphics[scale=0.45]{paperfigures/stormavs-md12.eps}
\caption{(a) Same as Figure \ref{Storms}(a) except with constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours. (b) Same as Figure \ref{Storms}(b) except for constrain that $\rho_{eq}$ stayed above $50$~amu/cm$^3$ for at least 12 hours.}
\label{Mspec}
\end{figure}
\clearpage

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/stormavs-d80.eps}
\rule[1ex]{5cm}{0.5pt}
\includegraphics[scale=0.45]{paperfigures/stormavs-m70.eps}
\caption{(a) Same as Figure \ref{Storms}(a) except with constraint that $D_{st}$ crossed below $-80$~nT (b) Same as Figure \ref{Storms}(b) except for constrain that $\rho_{eq}$ crossed above $70~amu/cm^3$.}
\label{Dspec}
\end{figure}

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/stormavs-diffden-10amu.eps}
\rule[1ex]{5cm}{0.5pt}
\includegraphics[scale=0.45]{paperfigures/stormavs-diffden-30percent.eps}
\caption{(a) diff($\rho_{eq}$) $>$ 10$\frac{amu}{hour}$. (b) diff($\rho_{eq}$) $>$ 30$\frac{\%}{hour}$}
\label{rhochange}
\end{figure}
\clearpage

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/stormavs-dst-nof107.eps}
\includegraphics[scale=0.45]{paperfigures/f107removed.eps}
\caption{$D_{st}$ events where $\rho_{eq}$ has $F_{10.7}$ dependence removed, and Difference between original $\rho_{eq}$ and $F_{10.7}$-less $\rho_{eq}$}
\label{nof107}
\end{figure}

\begin{figure}[htp]
\centering
\includegraphics[scale=0.45]{paperfigures/blockmedian.eps}
\includegraphics[scale=0.45]{paperfigures/blockmedian-prenoon.eps}
\caption{Comparing median values calculated as median of all 1-hour points in the 12 day block, and as the daily median of all hourly-mediated events. (b) Same, but only events starting between 0-12 UTC}
\label{blockmedian}
\end{figure}
\clearpage

\begin{figure}[htp]
\centering
\includegraphics[scale=0.9]{paperfigures/allstorms.eps}
\caption{Comparing all events to their median and $1\sigma$ range}
\label{allstorms}
\end{figure}
\clearpage

\begin{figure}[htp]
\centering
\includegraphics[scale=0.5]{paperfigures/nansbyhour.eps}
\includegraphics[scale=0.5]{paperfigures/nansbyhour_storm.eps}
\caption{(a) Number of NaN points per hour of observation in the total data set (b) $\rho_{eq}$ data availability relative to average event hour in events where $D_{st}<-80$~nT.}
\label{nanperhour}
\end{figure}
\clearpage

>>>>

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period and 1991 and found that the average ion mass density, $M=\rho/n_e$, correlated with geomagnetic activity, with lower $D_{st}$ values corresponding to slightly average ion mass densities.  They also found a relationship between the average ion mass and 1.5- and 3-day averages of $K_p$ were compared to $M$.  Most measurements occurred in the MLT range of 12:00 and 18:00 in the plasma trough, and CRRES had a nearly equatorial orbit.

\cite{Takahashi2010} developed a mass density dataset using measurements from the GOES satellites over the years 1980 through 1992, with most measurements in the range $L=6.8\pm0.2$. The equatorial mass density, $\rho_{eq}$ was estimated from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}(LR_E/R)$. The mass density $\rho$ was estimated using the the Alfven wave velocity relationship, $V_A=(\mu_0\rho)^{1/2}$, a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of an infinitely conducting ionosphere.

\cite{Takahashi2010} noted that downward spikes in the $D_{st}$ index coincide with with significant changes in $\rho_{eq}$ at an L-shell of 6.8~$R_E$. For five storms over a 20-day period, two had $\rho_{eq}$ spikes after the $D_{st}$ drop, two had $\rho_{eq}$ spikes before the drop, and one showed little change in $\rho_{eq}$.  A key result was that when daily-averaged measurements were considered, an enhancment in $\rho_{eq}$ appeared on the same day as minimum $D_{st}$, indicating the possibility of a solar wind control over $\rho_{eq}$.

The results of \cite{Takahashi2006} and \cite{Takahashi2010} are consistent with other results on changes in mass density in the plasma trough region.  \cite{Yao2008} studied the relationship between $D_{st}$ and ion number density for different ions in different regions (ring current and plasma sheet) and found a general correlation during each of the four storms considered.

\cite{Hamilton1988} ... \cite{Gallagher1988} ....

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset associated with \cite{Denton}, with data available from 1980 through 1991; all other parameters are from \cite{Reconstruction} over the time range of 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hourly range was used.  Fill values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple Geostationary Operational Environmental Satellites (GOES) satellites at the same time in the \cite{Denton} dataset, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data is shown in Figure~\ref{AllData}.

@@

\newpage
\footnotesize
\bibliographystyle{plainnat}
\bibliography{paper}

@@

\begin{figure}[htp!]
\centering
\subcaptionbox{\label{ccplot:a}}
{\includegraphics[scale=0.40]{paperfigures/F107MDAllData.eps}}
\subcaptionbox{\label{ccplot:b}}
{\includegraphics[scale=0.40]{paperfigures/ccplot.eps}}
\caption{(a) 27-day averages of $F_{10.7}$ and $\rho_{eq}$. (b) Correlation between $log(\rho_{eq})$ and $F_{10.7}$ using hour, day, and 27-day averaging intervals; compare to \cite{Takahashi2010} Fig.~14.}
\label{ccplot}
\end{figure}
\clearpage

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dst-50-tak.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-mass-tak.eps}
\caption{Events in the interval 1989-1991. Upper panel: $D_{st}$ events. Lower panel: $\rho_{eq}$ events.}
\label{DailyAverages}
\end{figure}

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dst-day.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-mass-day.eps}
\caption{Events in the interval 1983-1991.  Upper panel: $D_{st}$ events. Lower panel: $\rho_{eq}$ events.}
\label{DailyAveragesAllData}
\end{figure}
\clearpage

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dst.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-mass.eps}
\caption{(a) Average of solar wind and near-Earth measurements around the time $D_{st}$ crossed below $-50$~nT onset. (b) Same as (a) except around time intervals where mass density crossed above $40$~amu/cm$^3$.}
\label{Storms}
\end{figure}

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dd12.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-md12.eps}
\caption{(a) Same as Figure \ref{Storms}(a) except with constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours. (b) Same as Figure \ref{Storms}(b) except for constrain that $\rho_{eq}$ stayed above $50$~amu/cm$^3$ for at least 12 hours.}
\label{Mspec}
\end{figure}
\clearpage

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-d80.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-m70.eps}
\caption{(a) Same as Figure \ref{Storms}(a) except with constraint that $D_{st}$ crossed below $-80$~nT (b) Same as Figure \ref{Storms}(b) except for constrain that $\rho_{eq}$ crossed above $70~amu/cm^3$.}
\label{Dspec}
\end{figure}

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-diffden-10amu.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-diffden-30percent.eps}
\caption{(a) diff($\rho_{eq}$) $>$ 10$\frac{amu}{hour}$. (b) diff($\rho_{eq}$) $>$ 30$\frac{\%}{hour}$}
\label{rhochange}
\end{figure}
\clearpage

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dst-nof107.eps}
\includegraphics[scale=0.40]{paperfigures/f107removed.eps}
\caption{$D_{st}$ events where $\rho_{eq}$ has $F_{10.7}$ dependence removed, and Difference between original $\rho_{eq}$ and $F_{10.7}$-less $\rho_{eq}$}
\label{nof107}
\end{figure}

\begin{figure}[tp]
\centering
\includegraphics[scale=0.40]{paperfigures/blockmedian.eps}
\includegraphics[scale=0.40]{paperfigures/blockmedian-prenoon.eps}
\caption{Comparing median values calculated as median of all 1-hour points in the 12 day block, and as the daily median of all hourly-mediated events. (b) Same, but only events starting between 0-12 UTC}
\label{blockmedian}
\end{figure}
\clearpage

\begin{figure}[tp]
\centering
\includegraphics[scale=0.9]{paperfigures/allstorms.eps}
\caption{Comparing all events to their median and $1\sigma$ range}
\label{allstorms}
\end{figure}
\clearpage

\begin{figure}[tp]
\centering
\includegraphics[scale=0.5]{paperfigures/nansbyhour.eps}
\includegraphics[scale=0.5]{paperfigures/nansbyhour_storm.eps}
\caption{(a) Number of NaN points per hour of observation in the total data set (b) $\rho_{eq}$ data availability relative to average event hour in events where $D_{st}<-80$~nT.}
\label{nanperhour}
\end{figure}
\clearpage
